Untitled

### 2022-04-08 Papers/Polymethylation_scores/06_discussion.Rmd
coauthor feedback april 4

Our findings are consistent with the previous literature in different tissues and at different ages on associations between prenatal maternal smoking and DNA methylation. We replicated top hits from the seminal epigenome wide association meta-analysis of prenatal smoke exposure and DNA methylation of cord blood from newborns including *AHRR*:cg05575921, *GFl1*:cg14179389 and *MYO1G1*:cg22132788. Consistent with a meta-analysis of peripheral blood from children and adults over 16 years of age, the association at *MYO1G1*:cg22132788 was larger than in newborn cord blood [@Joubert:2016ue; @wiklund2019]. Again consistent with a meta-analysis of peripheral blood from children and adults over 16 years of age, the associations at *AHRR*:cg05575921 and *GFl1*:cg14179389 were smaller than in newborn cord blood [@Joubert:2016ue; @wiklund2019]. There are no previous studies of prenatal maternal smoking and saliva DNA methylation. However, immune cell types are present in both blood and saliva, and saliva samples in our study were primarily composed of immune cells. Additionally, fewer than 4% of DNA methylation sites in blood and saliva samples displayed significantly different methylation values ($|\Delta \beta| >0.2$ and $p_{adj}< 0.001$) when measured on the 450K chip in samples of adults [@Langie2016] and 11-year old children [@Langie2017]. Our results suggest salivary DNA methylation can be used as an alternative to peripheral blood DNA methylation for biomarkers of prenatal maternal smoking. A saliva-based DNA methylation biomarker may be particularly useful in cases where blood collection, processing, and storage is not possible. Saliva can often be collected directly by participants and saliva sample collection kits have stabilizers allowing room temperature storage prior to DNA methylation measurement.

We advance the prenatal smoking - DNA methylation literature by evaluating the generalizability of association between prenatal maternal smoking and DNA methylation across genetic ancestry groups. The portability of polygenic risk scores across genetic ancestries is a complex research area [@Martin2019] and evaluating the portability of epigenetic summary measures has been identified as a key area for evaluation [@Huls2020]. Our results suggest polymethylation scores for prenatal maternal smoking are portable across genetic ancestry groups. At age 9, the effect estimates for prenatal maternal smoking on the polymethylation score in the  European and African genetic ancestry samples were was within `r abs(coefOnly(local_models, "childteen=='Age 9' & predictors=='base model' & str_detect(outcome, 'cell-type') & ancestry=='European ancestry'", 'smkPreg_binaryYes', 2)-coefOnly(local_models, "childteen=='Age 9' & predictors=='base model' & str_detect(outcome, 'cell-type') & ancestry=='African ancestry'", 'smkPreg_binaryYes', 2))` standard deviation and were both significant at an alpha of 0.05. This finding is in line with a recent epigenome-wide association study in which children of both black and non-Black identifying mothers exhibited similar directions of associations between prenatal smoke exposure and peripheral blood methylation at 38 DNA methylation sites [@Xu2021]. In our analysis, the association between prenatal smoke exposure and polymethylation scores was attenuated in children of admixed genetic ancestry (primarily of Hispanic ethnicity) at both ages and in children of European genetic ancestry at age 15. This could reflect differences in DNA methylation by genetic sequence at prenatal smoke sensitive DNA methylation sites [@Gonseth2016]. While genetic ancestry is distinct from racial and ethnic social and cultural constructs [@borrell2021], it can correlate with different social and environmental exposures. Differences in other social and environmental exposures could also contribute to the attenuation observed in the admixed genetic ancestry group.

@@

While polymethylation scores significantly improved classification of prenatal smoke exposure over a base model, using *AHRR*:cgg05575921 methylation as a classifier did not (percentage point increase in AUC: `r rct1%>%filter(childteen=='Age 9' & str_detect(modeltype, 'AHRR'))%>%pull(5)`) . Prenatal smoke exposed newborns consistently exhibit lower DNA methylation at *AHRR*:cg05575291 than unexposed newborns [@Joubert:2016ue]. However, the association between prenatal smoke exposure and *AHRR*:cg05575291 methylation did not persist in peripheral blood samples from adult women [@Richmond:2018tu]. Additionally, a previous longitudinal analysis of children's blood samples from birth to 17 years of age in the ALSPAC cohort found that the difference in DNA methylation at *AHRR*:cg05575291 between prenatal smoke exposed and unexposed children decreased from birth to 17 years of age. This previous analysis also reported that differences in DNA methylation at *AHRR*:cg05575291 in adolescence may be partially mediated by postnatal own smoking, and we excluded children who reported postnatal own-smoking [@Richmond:2018tu]. As the accuracy of *AHRR*:cg05575921 and other single CpG site biomarkers may be influenced by time-since-exposure and new environmental exposures, incorporating information across multiple sites of DNA methylation may yield a a more robust biomarker.

Additionally, we found that prenatal maternal smoke exposure was not associated with global DNA methylation or DNA methylation clocks, underlining the specificity of the DNA methylation signature of prenatal smoke exposure. DNA methylation clocks attempt to measure aspects of biological aging and are trained in specific populations. We note that the pediatric DNA methylation clock, which was specifically trained for use in pediatric populations, accurately recapitulated our subjects chronological ages. However, the GRIM age clock, which was trained in adults to measure proximity to mortality, overestimated chronological ages in our population by 10-15 years. In an analysis of 4 pediatric cohorts of children under 20 years of age, GRIM age estimates were 30-35 years greater than chronological in buccal cell samples and 10-15 years greater than chronological age in blood samples [@okazaki2021]. GRIM age was neither trained nor validated in a pediatric population [@lu2019], which may explain this overestimation.

Our analysis is not without its limitations. We used maternal self-report of prenatal maternal smoke exposure, as serum cotinine levels were not available. As mothers may be reluctant to report smoking during pregnancy, some of our children may be misclassified as unexposed to prenatal smoke. We would expect this to bias our results towards the null. Analyses of prenatal exposures and postnatal outcomes are conditioned on live birth and survival or enrollment until the point of outcome measurement. Because prenatal maternal smoking is associated with miscarriage [@pineles2014], this could result in selection bias if an unmeasured variable is associated with both live birth and the outcome. This selection bias can create a downward bias in the effect estimate. This could produce a spurious protective effect (which we did not observe), or, in the case of a true positive association between exposure and outcome, will attenuate the association toward the null[@leung2021; @liew2015]. Additionally, while we controlled for postnatal secondhand smoke exposure and excluded children who reported any own-smoking, we cannot exclude the possibility some of the observed effects are mediated through these alternative sources of smoke exposure. Personal smoking behavior among children in our cohort was relatively rare, and we excluded those participants from our analysis to focus on the prenatal exposure window. Future studies may focus on enriched risk samples to examine the joint effect of prenatal smoke exposure and personal smoking.

>>>>

Our findings are consistent with the previous literature in different tissues and at different ages on associations between prenatal maternal smoking and DNA methylation. We replicated top hits from the seminal epigenome wide association meta-analysis of prenatal smoke exposure and DNA methylation of cord blood from newborns including *AHRR*: cg05575921, *GFl1*: cg14179389 and *MYO1G1*: cg22132788. Consistent with a meta-analysis of peripheral blood from children and adults over 16 years of age, the association at *MYO1G1*: cg22132788 was larger than in newborn cord blood [@Joubert:2016ue; @wiklund2019]. Again, consistent with a meta-analysis of peripheral blood from children and adults over 16 years of age, the associations at *AHRR*: cg05575921 and *GFl1*: cg14179389 were smaller than in newborn cord blood [@Joubert:2016ue; @wiklund2019]. There are no previous studies of prenatal maternal smoking and saliva DNA methylation. However, immune cell types are present in both blood and saliva, and saliva samples in our study were primarily composed of immune cells. Additionally, fewer than 4% of DNA methylation sites in blood and saliva samples displayed significantly different methylation values ($|\Delta \beta| >0.2$ and $p_{adj}< 0.001$) when measured on the 450K chip in samples of adults [@Langie2016] and 11-year old children [@Langie2017]. Our results suggest salivary DNA methylation can be used as an alternative to peripheral blood DNA methylation for biomarkers of prenatal maternal smoking. A saliva-based DNA methylation biomarker may be particularly useful in cases where blood collection, processing, and storage is not possible. Saliva can often be collected directly by participants and saliva sample collection kits have stabilizers allowing room temperature storage prior to DNA methylation measurement.

We advance the prenatal smoking - DNA methylation literature by evaluating the generalizability of association between prenatal maternal smoking and DNA methylation across genetic ancestry groups. The portability of polygenic risk scores across genetic ancestries is a complex research area [@Martin2019] and evaluating the portability of epigenetic summary measures has been identified as a key area for evaluation [@Huls2020]. Our results suggest polymethylation scores for prenatal maternal smoking are portable across genetic ancestry groups. At age 9, the effect estimates for prenatal maternal smoking on the polymethylation score in the European and African genetic ancestry samples were was within `r abs(coefOnly(local_models, "childteen=='Age 9' & predictors=='base model' & str_detect(outcome, 'cell-type') & ancestry=='European ancestry'", 'smkPreg_binaryYes', 2)-coefOnly(local_models, "childteen=='Age 9' & predictors=='base model' & str_detect(outcome, 'cell-type') & ancestry=='African ancestry'", 'smkPreg_binaryYes', 2))` standard deviation and were both significant at an alpha of 0.05. This finding is in line with a recent epigenome-wide association study in which children of both black and non-black identifying mothers exhibited similar directions of associations between prenatal smoke exposure and peripheral blood methylation at 38 DNA methylation sites [@Xu2021]. In our analysis, the association between prenatal smoke exposure and polymethylation scores was attenuated in children of admixed genetic ancestry (primarily of Hispanic ethnicity) at both ages and in children of European genetic ancestry at age 15. This could reflect differences in DNA methylation by genetic sequence at prenatal smoke sensitive DNA methylation sites [@Gonseth2016]. While genetic ancestry is distinct from racial and ethnic social and cultural constructs [@borrell2021], it can correlate with different social and environmental exposures. Differences in other social and environmental exposures could also contribute to the attenuation observed in the admixed genetic ancestry group.

@@

While polymethylation scores significantly improved classification of prenatal smoke exposure over a base model, using *AHRR*: cg05575921 methylation as a classifier did not (percentage point increase in AUC: `r rct1%>%filter(childteen=='Age 9' & str_detect(modeltype, 'AHRR'))%>%pull(5)`)). Prenatal smoke exposed newborns consistently exhibit lower DNA methylation at *AHRR*: cg05575291 than unexposed newborns [@Joubert:2016ue]. However, the association between prenatal smoke exposure and *AHRR*: cg05575291 methylation did not persist in peripheral blood samples from adult women [@Richmond:2018tu]. Additionally, a previous longitudinal analysis of children's blood samples from birth to 17 years of age in the ALSPAC cohort found that the difference in DNA methylation at *AHRR*: cg05575291 between prenatal smoke exposed and unexposed children decreased from birth to 17 years of age. This previous analysis also reported that differences in DNA methylation at *AHRR*: cg05575291 in adolescence may be partially mediated by postnatal own smoking, and we excluded children who reported postnatal own-smoking [@Richmond:2018tu]. As the accuracy of *AHRR*: cg05575921 and other single CpG site biomarkers may be influenced by time-since-exposure and new environmental exposures, incorporating information across multiple sites of DNA methylation may yield a a more robust biomarker.

Additionally, we found that prenatal maternal smoke exposure was not associated with global DNA methylation or a pediatric saliva and the GrimAge DNA methylation clocks, underlining the specificity of the DNA methylation signature of prenatal smoke exposure. DNA methylation clocks attempt to measure aspects of biological aging and are trained in specific populations. We note that the pediatric DNA methylation clock, which was specifically trained for use in pediatric populations, accurately recapitulated our subjects chronological ages. However, the GRIM age clock, which was trained in adults to measure proximity to mortality, overestimated chronological ages in our population by 10-15 years. In an analysis of four pediatric cohorts of children under 20 years of age, GRIM age estimates were 30-35 years greater than chronological in buccal cell samples and 10-15 years greater than chronological age in blood samples [@okazaki2021]. GRIM age was neither trained nor validated in a pediatric population [@lu2019], which may explain this overestimation.

Our analysis is not without its limitations. We used maternal self-report of prenatal maternal smoke exposure, as serum cotinine levels were not available. As mothers may be reluctant to report smoking during pregnancy, some of our children may be misclassified as unexposed to prenatal smoke. We would expect this to bias our results towards the null. Analyses of prenatal exposures and postnatal outcomes are conditioned on live birth and survival or enrollment until the point of outcome measurement. Because prenatal maternal smoking is associated with miscarriage [@pineles2014], this could result in selection bias if an unmeasured variable is associated with both live birth and the outcome. This selection bias can create a downward bias in the effect estimate. This could produce a spurious protective effect (which we did not observe), or, in the case of a true positive association between exposure and outcome, will attenuate the association toward the null [@leung2021; @liew2015]. Additionally, while we controlled for postnatal secondhand smoke exposure and excluded children who reported any own-smoking, we cannot exclude the possibility some of the observed effects are mediated through these alternative sources of smoke exposure. Personal smoking behavior among children in our cohort was relatively rare, and we excluded those participants from our analysis to focus on the prenatal exposure window. Future studies may focus on enriched risk samples to examine the joint effect of prenatal smoke exposure and personal smoking.

### 2022-04-08 Papers/Polymethylation_scores/08_methods.Rmd
coauthor feedback april 4

The Fragile Families and Child Wellbeing Study is a birth cohort of nearly 5,000 children born in 20 cities in the United States between 1998 and 2000 [@Reichman:2001wv]. Participants were selected at delivery using a three-stage stratified random sample design which oversampled unmarried mothers by a ratio of 3:1 [@Reichman:2001wv]. Participants were excluded at the baseline wave on the following criteria: the parents who planned to place the child for adoption, the father was deceased, the parents did not speak English or Spanish well enough to be interviewed, or the mother or baby were too ill to complete the interview, or the baby died before the interview could take place. Children were followed longitudinally with assessments at ages 1, 3, 5, 9 and 15; an age 22 follow up is ongoing. Across certain child ages, assessments included medical record extraction, biosample collection, in-home assessments, and surveys of the mother, father, primary caregiver, teacher and child. At ages nine and fifteen a saliva sample was taken from the child [@Reichman:2001wv]. A subsample of the Fragile Families cohort was selected for saliva DNA and DNA methylation processing [@Reichman:2001wv]. To be eligible for the DNA methylation assessment children had to have participated and given saliva at age 9 and 15; after which a random sample of these eligible participants was then selected. One exception to this rule is that all participants in the Study of Adolescent Neurodevelopment (SAND) were assayed even if they only provided a sample at one time-point. All analyses account for this oversample of these participants.

@@

Prenatal maternal substance use variables were derived from maternal self-report questionnaire data at baseline (child's birth).

At baseline, mother's answered categorical questions about their prenatal smoking, drug and alcohol use. For maternal prenatal smoking, mother's were asked: "During your pregnancy, how many cigarettes did you smoke?". Responses were categorical: 2 or more packs a day, 1 or more but less than 2 packs per day, less than 1 pack a day or none. Few participants reported smoking a pack or more a day, thus for our primary exposure variable we created a dichotomous any versus no prenatal smoke exposure variable.

@@

In sensitivity analyses, we controlled for postnatal second hand smoke exposure. When the child was 1, 5, 9 and 15 years of age, primary caregivers responded to questions about smoking in the household. To encapsulate early life postnatal secondhand smoke exposure, we created a binary variable for postnatal exposure at ages 1 or 5. To encapsulate recent postnatal smoke exposure to the age 9 and 15 interview where DNA methylation was collected, we used a categorical variable for packs per day (no smoking, less than one pack/day, one or more packs/day) in the month prior to each interview.

To ensure we measured the direct effect of prenatal exposure to cigarette smoke, rather than the effect of personal postnatal smoking we excluded children who reported own smoking. At ages 9 children were asked if they had ever smoked a cigarette or used tobacco (yes/no) and at age 15 they were asked if they had ever smoked an entire cigarette (yes/no).

@@

Child genetic ancestry was calculated. Grants R01 HD36916, R01 HD073352, and R01 HD076592 provided support for the collection, assay, and analysis of the genetic data in the cohort. Principal components (PC) of child genetic ancestry were calculated from genetic data using Illumina PsychChip_v1-1 with child saliva samples. Genetic ancestry was assigned by comparing PC loadings to 1000 Genomes super population clusters. Samples with PC1 \> 0.018 and PC2 \> -0.0075 were assigned to European ancestry. Samples with PC1 \< -0.005 and PC2 \> 0.007+0.75(PC1) were assigned to African ancestry. Samples with PC1 \> 0.018 and -0.055 \< PC2 \< 0.025 were assigned to Admixed ancestry [@FFPGS2021]. Samples assigned to Admixed ancestry were primarily from individuals whose mothers self-reported Hispanic ethnicity (`r modeldata%>%group_by(ancestry, cm1ethrace)%>%summarise(n=n())%>%ungroup()%>%group_by(ancestry)%>%mutate(sum=sum(n), percent=round(100*(n/sum), 0))%>%filter(ancestry=='Admixed ancestry - Latin heritage' & cm1ethrace=='3 hispanic')%>%pull(percent)`% of samples), so we label this group Admixed - Latin ancestry.

@@

Salivary samples from the children were collected at ages nine and fifteen using the Oragene•DNA sample collection kit (DNA Genotek Inc., Ontario). Saliva DNA was extracted manually using DNA Genotek's purification protocol using prepIT L2P. DNA was bisulfite treated and cleaned using the EZ DNA Methylation kit (Zymo Research, California). Samples were randomized and plated across slides by demographic characteristics. Saliva DNA methylation was measured using the Illumina HumanMethylation 450k BeadArray [@Sandoval2011] and imaged using the Illumina iScan system. All samples were run in a single batch to minimize technical variability.

DNA methylation image data were processed in R statistical software (3.5) using the minfi package [@Aryee2014]. We used the February 2022 data freeze of the Fragile Families and Child Wellbeing DNA methylation data. The red and green image pairs (n=1811) were read into R and the minfi *preprocessNoob* function was used to normalize dye bias and apply background correction. Further quality control was applied using the ewastools packages [@Heiss2018]. We dropped samples using the following criteria: if \>10% of DNA methylation sites had detection p-value \>0.01 (n=`r probe_fail`), if there was sex discordance between DNA methylation predicted sex and recorded sex (n=`r sex_mismatch`), if samples had outlier methylation values (n=`r meth_outlier`) or if two sequential samples from the same individual exhibited genetic discordance between visits (n=`r genetic_discordance`) (Figure 1). Technical replicates were removed (n=`r tech_rep`). DNA methylation sites were removed if they had detection p-value \>0.01 in 5% of samples (n=33376) [@Chen2013]. Relative proportions of immune and epithelial cell types were estimated from DNA methylation measures using a childhood saliva reference panel [@Middleton2021].

@@

To compare the utility of polymethylation scores relative to single site biomarkers, we selected the top DNA metyhlation site from the Joubert meta-analysis, *AHRR:* cg05575921. As a sensitivity analyses, we also selected four other *a priori* DNA methylation sites which were identified from meta-analyses of prenatal maternal smoking and DNA methylation in children's cord and peripheral blood samples: *GFI1*:cg14179389, *CYP1A1*: cg05549655, *MYO1G*: cg22132788, and *MYO1G*:cg04180046 [@Joubert:2016ue; @wiklund2019].

@@

We excluded samples missing any data on the following variables: maternal prenatal smoking, maternal prenatal alcohol use, maternal prenatal other drug use, maternal income to poverty ratio, postnatal secondhand smoke exposure variables, child sex, or child age. We also excluded samples that were missing genetic data. Age 9 samples from children who reported ever smoking a cigarette or using tobacco age 9 or were missing own-smoking data were excluded. Age 15 samples from children who reported smoking a cigarette at age 9 or at age 15 or were missing own-smoking data were excluded. Children who were missing a response to the question at age 9 but answered that they had never smoked a whole cigarette at age 15 were kept in the sample.

@@

We described the distribution of continuous variables using the mean and standard deviation. We described the distribution of categorical variables using percents. We compared the distribution of included to excluded samples using the Welch's t-test for continuous variables and the Chi-Square test for categorical variables. We report the correlation between continuous measures of DNA methylation (including polymethylation scores, single sites, mean global DNA methylation and estimated cell type proportions) using the Pearson correlation coefficient . We report the correlation of these DNA methylation measures between visits using the Pearson correlation coefficient and scatter plots. We compared the distribution of variables by exposure status to prenatal maternal smoking using the Welch's t-test for continuous variables and the Chi-Square test for categorical variables. For select continuous DNA methylation measures we compare the distribution across prenatal smoke exposure status using violin plots with captive box plots.

@@

To assess the association between our primary exposure variable of prenatal maternal smoking and salivary DNA methylation, we performed age-stratified multivariable linear regressions with the primary polymethlation score as our outcome. To facilitate interpretation, we z-score standardized the polymethylation score. We adjusted for an indicator variable for maternal residence in Chicago, Detroit or Toledo (over sampled cities), maternal income to poverty ratio at baseline, child sex, child age, plate from DNA methylation processing, estimated immune cell proportion from DNA methylation data, and the first two components of genetic ancestry from principal component analysis. While child sex, child age, plate from DNA methylation processing and immune cell proportions are not confounders (as they cannot casually effect prenatal maternal smoke exposure), these variables can strongly effect DNA methylation and so were adjusted for as precision variables. We reported the effect estimate for polymethylation score term and the 95% confidence interval (CI). We visualized our findings using forest plots.

@@

To compare the utility of the polymethylation score to a single site biomarkers, we used percent methylation at AHRR:cg05575921 as the outcome in the models described above. To assess the specificity of the impact of prenatal maternal smoking on DNA methylation, we used mean percent global DNA methylation and the pediatric DNA methylation clock as outcome variables.

@@

To assess the robustness of the association to choice of polymethylation score, we performed the same regression analyses described above with z-score standardized sensitivity polymethylation scores as the outcome. Similarly, we replaced the outcome variable of *AHRR:*cg05575921 with other *a priori* selected DNA methylation sites in sensitivity analyses. We plotted the estimates and 95% CI for each DNA methylation site along with the previously reported estimates from the literature using forest plots [@Joubert:2016ue; @wiklund2019]

@@

To evaluate the accuracy of the DNA methylation summary measures as biomarkers of prenatal maternal smoking, we used a receiver operating curve. First, we regressed exposure to prenatal maternal smoking (outcome) against the DNA methylation summary measures in individual logistic regressions, while adjusting for the base model variables listed above. Next, we calculated receiver operating curves (ROC) and area under the curves (AUCs) using the function roc from the R library pROC version 1.18.0 [@pROC2011]. We compared ROC curves and AUCs from models with only base model variables to those which included base model variables and the methylation summary measure using the Delong method and the function roc.test from the R library pROC version 1.18.0. This is appropriate as the base model variable only model is nested within the base model variable + methylation summary measures model.

>>>>

The Fragile Families and Child Wellbeing Study is a birth cohort of nearly 5,000 children born in 20 cities in the United States between 1998 and 2000 [@Reichman:2001wv]. Participants were selected at delivery using a three-stage stratified random sample design which oversampled unmarried mothers by a ratio of 3:1 [@Reichman:2001wv]. Participants were excluded at the baseline wave on the following criteria: the parents who planned to place the child for adoption, the father was deceased, the parents did not speak English or Spanish well enough to be interviewed, the mother or baby were too ill to complete the interview, or the baby died before the interview could take place. Children were followed longitudinally with assessments at ages 1, 3, 5, 9 and 15; an age 22 follow up is ongoing. Across certain child ages, assessments included medical record extraction, biosample collection, in-home assessments, and surveys of the mother, father, primary caregiver, teacher and child. At ages nine and fifteen a saliva sample was taken from the child [@Reichman:2001wv]. A subsample of the Fragile Families cohort was selected for saliva DNA and DNA methylation processing [@Reichman:2001wv]. To be eligible for the DNA methylation assessment children had to have participated and given saliva at age 9 and 15; after which a random sample of these eligible participants was then selected. One exception to this rule is that all participants in the Study of Adolescent Neurodevelopment (SAND) were assayed even if they only provided a sample at one time-point. All analyses account for oversampling of these participants.

@@

Prenatal maternal substance use variables were derived from maternal self-report questionnaire data at baseline (child's birth). At baseline, mothers answered categorical questions about their prenatal smoking, drug and alcohol use. For maternal prenatal smoking, mother's were asked: "During your pregnancy, how many cigarettes did you smoke?". Responses were categorical: 2 or more packs a day, 1 or more but less than 2 packs per day, less than 1 pack a day or none. Few participants reported smoking a pack or more a day, thus for our primary exposure variable we created a dichotomous any versus no prenatal smoke exposure variable.

@@

In sensitivity analyses, we controlled for postnatal second-hand smoke exposure. When the child was 1, 5, 9 and 15 years of age, primary caregivers responded to questions about smoking in the household. To encapsulate early life postnatal secondhand smoke exposure, we created a binary variable for postnatal exposure at ages 1 and/or 5. To encapsulate recent postnatal smoke exposure to the age 9 and 15 interview where DNA methylation was collected, we used a categorical variable for packs per day (no smoking, less than one pack/day, one or more packs/day) in the month prior to each interview.

To ensure we measured the direct effect of prenatal exposure to cigarette smoke, rather than the effect of personal postnatal smoking we excluded children who reported own smoking. At age 9 children were asked if they had ever smoked a cigarette or used tobacco (yes/no), and at age 15 they were asked if they had ever smoked an entire cigarette (yes/no).

@@

Child genetic ancestry was calculated. Principal components (PC) of child genetic ancestry were calculated from genetic data using Illumina PsychChip with child saliva samples. Genetic ancestry was assigned by comparing PC loadings to 1000 Genomes super population clusters. Samples with PC1 \> 0.018 and PC2 \> -0.0075 were assigned to European ancestry. Samples with PC1 \< -0.005 and PC2 \> 0.007+0.75(PC1) were assigned to African ancestry. Samples with PC1 \> 0.018 and -0.055 \< PC2 \< 0.025 were assigned to Admixed ancestry [@FFPGS2021]. Samples assigned to Admixed ancestry were primarily from individuals whose mothers self-reported Hispanic ethnicity (`r modeldata%>%group_by(ancestry, cm1ethrace)%>%summarise(n=n())%>%ungroup()%>%group_by(ancestry)%>%mutate(sum=sum(n), percent=round(100*(n/sum), 0))%>%filter(ancestry=='Admixed ancestry - Latin heritage' & cm1ethrace=='3 hispanic')%>%pull(percent)`% of samples), so we label this group Admixed - Latin ancestry.

@@

Salivary samples from the children were collected at ages 9 and 15 using the Oragene DNA sample collection kit (DNA Genotek Inc., Ontario). Saliva DNA was extracted manually using DNA Genotek's purification protocol using prepIT L2P. DNA was bisulfite treated and cleaned using the EZ DNA Methylation kit (Zymo Research, California). Age 9 and 15 samples were run on the same slide to minimize technical variation, and otherwise samples were randomized. Saliva DNA methylation was measured using the Illumina HumanMethylation 450k BeadArray [@Sandoval2011] and imaged using the Illumina iScan system. The bead arrays were processed at the Genome Sciences Core Facility at the Pennsylvania University College of Medicine (Hershey, PA).

DNA methylation image data were processed in R statistical software (4.1) using the Enmix package [@Xu2021a]. We used the February 2022 data freeze of the Fragile Families and Child Wellbeing DNA methylation data. The red and green image pairs (n=1811) were read into R and the ENmix preprocessENmix and rcp functions were used to normalize dye bias, apply background correction and adjust for probe-type bias. Further quality control was applied using the ewastools packages [@Heiss2018]. We dropped samples using the following criteria: if \>10% of DNA methylation sites had detection p-value \>0.01 (n=`r probe_fail`), if there was sex discordance between DNA methylation predicted sex and recorded sex (n=`r sex_mismatch`), if samples had outlier methylation values (n=`r meth_outlier`) or if two sequential samples from the same individual exhibited genetic discordance between visits (n=`r genetic_discordance`) (Figure 1). Technical replicates were removed (n=`r tech_rep`). DNA methylation sites were removed if they had detection p-value \>0.01 in 5% of samples (n=33376) [@Chen2013]. Relative proportions of immune and epithelial cell types were estimated from DNA methylation measures using a childhood saliva reference panel [@Middleton2021].

@@

To compare the utility of polymethylation scores relative to single site biomarkers, we selected the top DNA methylation site from the 2016 Joubert meta-analysis, *AHRR:* cg05575921. As a sensitivity analyses, we also selected four other *a priori* DNA methylation sites which were identified from meta-analyses of prenatal maternal smoking and DNA methylation in children's cord and peripheral blood samples: *GFI1*: cg14179389, *CYP1A1*: cg05549655, *MYO1G*: cg22132788, and *MYO1G*: cg04180046 [@Joubert:2016ue; @wiklund2019].

@@

We excluded samples missing any data on the following variables: maternal prenatal smoking, maternal prenatal alcohol use, maternal prenatal other drug use, maternal income to poverty ratio, postnatal secondhand smoke exposure variables, child sex, or child age. We also excluded samples that were missing genetic data. Age 9 samples from children who reported ever smoking a cigarette or using tobacco at age 9 or were missing own-smoking data were excluded. Age 15 samples from children who reported smoking a cigarette at age 9 or at age 15 or were missing own-smoking data were excluded. Children who were missing a response to the question at age 9 but answered that they had never smoked a whole cigarette at age 15 were kept in the sample.

@@

We described the distribution of continuous variables using the mean and standard deviation. We described the distribution of categorical variables using percents. We compared the distribution of included to excluded samples using the Welch's t-test for continuous variables and the Chi-Square test for categorical variables. We report the correlation between continuous measures of DNA methylation (including polymethylation scores, single sites, mean global DNA methylation and estimated cell type proportions) using the Pearson correlation coefficient. We report the correlation of these DNA methylation measures between visits using the Pearson correlation coefficient and scatter plots. We compared the distribution of variables by exposure status to prenatal maternal smoking using the Welch's t-test for continuous variables and the Chi-Square test for categorical variables. For select continuous DNA methylation measures we compare the distribution across prenatal smoke exposure status using violin plots with captive box plots.

@@

To assess the association between our primary exposure variable of prenatal maternal smoking and salivary DNA methylation, we performed age-stratified, multivariable linear regressions with the primary polymethlation score as our outcome. To facilitate interpretation, we z-score standardized the polymethylation score. We adjusted for an indicator variable for maternal residence in one of the oversampled cities (Chicago, Detroit or Toledo), maternal income to poverty ratio at baseline, child sex, child age, plate from DNA methylation processing, estimated immune cell proportion from DNA methylation data, and the first two components of genetic ancestry from principal component analysis. While child sex, child age, plate from DNA methylation processing and immune cell proportions are not confounders (as they cannot causally effect prenatal maternal smoke exposure), these variables can strongly effect DNA methylation and so were adjusted for as precision variables. We reported the effect estimate for polymethylation score term and the 95% confidence interval (CI). We visualized our findings using forest plots.

@@

To compare the utility of the polymethylation score to a single site biomarker, we used percent methylation at AHRR: cg05575921 as the outcome in the models described above. To assess the specificity of the impact of prenatal maternal smoking on DNA methylation, we used mean percent global DNA methylation and the pediatric DNA methylation clock as outcome variables.

@@

To assess the robustness of the association to choice of polymethylation score, we performed the same regression analyses described above with z-score standardized sensitivity polymethylation scores as the outcome. Similarly, we replaced the outcome variable of *AHRR:* cg05575921 with other *a priori* selected DNA methylation sites in sensitivity analyses. We plotted the estimates and 95% CI for each DNA methylation site along with the previously reported estimates from the literature using forest plots [@Joubert:2016ue; @wiklund2019]

@@

To evaluate the accuracy of the DNA methylation summary measures as biomarkers of prenatal maternal smoking, we used a receiver operating curve. First, we regressed exposure to prenatal maternal smoking (outcome) against the DNA methylation summary measures in individual logistic regressions, while adjusting for the base model variables from regression analyses. The base model variables included indicator variable for maternal residence in one of the oversampled cities (Chicago, Detroit or Toledo), maternal income to poverty ratio at baseline, child sex, child age, plate from DNA methylation processing, estimated immune cell proportion from DNA methylation data, and the first two components of genetic ancestry from principal component analysis. Next, we calculated receiver operating curves (ROC) and area under the curves (AUCs) using the function *roc* from the R library pROC version 1.18.0 [@pROC2011]. We compared ROC curves and AUCs from models with only base model variables to those which included base model variables and the methylation summary measure using the Delong method and the function *roc.test* from the R library pROC version 1.18.0. This is appropriate as the base model variable only model is nested within the base model variable + methylation summary measures model.

### 2022-04-08 Papers/Polymethylation_scores/98_declarations.Rmd
coauthor feedback april 4

Many of the variables used in this analysis are publicly available in the Fragile Families dataset. Some of the data used in this analysis, including genetic and epigenetic data, is restricted use but available to researchers upon reasonable application.

@@

This research was made possible through several grants (Fragile Families Core Data Collection: R01 HD036916, R25 AG053227, genetic data processing R01 HD36916, R01 HD073352, and R01 HD076592). FB was supported by the National Institutes of Health National Institute of Dental and Craniofacial Research (F31 DE029992). EB was supported by (R01 AG055406, R01 AG067592, R01 MD011716).

>>>>

Many of the variables used in this analysis are publicly available in the Fragile Families and Child Wellbeing dataset at http://opr.princeton.edu/archive/restricted/Default.aspx. The genetic and epigenetic data is restricted use but available to researchers upon standard application.

@@

This research was made possible through several grants (Fragile Families Core Data Collection: R01 HD036916, , genetic data processing R01 HD36916, R01 HD073352, R01 MD011716, and R01 HD076592). FB was supported by the National Institutes of Health National Institute of Dental and Craniofacial Research (F31 DE029992) and R25 AG053227. EB was supported by R01 AG055406, R01 AG067592, and R01 MD011716. LS, DN, and JF were supported by R01 HD076592, R01 MD011716. CM was supported by R01 AG07107, R01 MD011716, R01 HD076592.
